The five pillars of computational reproducibility: Bioinformatics and beyond

### 2023-08-17 manuscript/5pillars.Rmd
progress

For example, in 2006 an article entitled "Genomic signatures to guide the use of chemotherapeutics"
generated a great deal of interest as it was an early application of high throughput transcriptomics
in the prediction of individual patient responses to different chemotherapies @Potti2006-yk.
After observing some unusual features of the patient group, Baggerly and Coombes attempted
reproduction of some of the key findings @Baggerly2010-nr.
Without the help of scripted workflows to guide the re-analysis, the team used forensic bioinformatic
techniques to piece together how the dataset was originally analysed.
Their investigations found a litany of issues.
Firstly, the labeling of patients in the test set as "responders" or "non-responders" had been reversed
in the original analysis.
Secondly, some of the patients were included more than once (some up to four times) in the analysis,
likely to cause major distortions in results.
Confusingly, some of the reused data has inconsistent grouping; ie: some of the samples are labeled
both sensitive and resistant.
Additional errors include two cases where results (charts) were ascribed to the wrong drug.
Baggerly and Coombes highlight that such mistakes can inadvertently occur when conducting unscripted
data analysis such as using spreadsheets, and such problems can be obscured by a lack of documentation.
The article underwent two corrigenda, but was untimately retracted in 2011, as the authors were not
able to reproduce the findings themselves due to "corruption of several validation data sets"
@Potti2011-ut.
As the array findings were the basis for clinical trials where patients were allocated to treatments,
the flawed data analysis may have harmed patients given the wrong drug in the
period 2007-2010.
In 2010, Duke University terminated the trials and suspended the lead author, Dr Anil Potti, who
later resigned.
Duke was served eight lawsuits by families of affected patients seeking compensation for exposure to
harmful and unnecessary chemotherapy, which were settled out of court @Dyer2015-my.
This worst-case scenario emphasizes that computational reproducibility is crucial for
translating bioinformatics research into real-world outcomes.

@@

A number of guides recommending enhanced computational reproducibility practices have been developed
[ @Tan2010-is; @Peng2011-ia; @Sandve2013-lk; @Piccolo2016-ji; @Lewis2016-qh;
@Gruning2018-sg; @Akalin2018-eg; @Brito2020-ld; @Peikert2021-eq;
@Rodrigues2023-fw; @DuPre2022-bk].
Inspired by these principles, this article presents the
*five pillars of reproducible computational research* (**Figure 1**).
There are clear parallels with the established three pillars of the open science framework,
(open data, code and papers) [@Gorgolewski2016-hv].
Here, the emphasis is on practical reproducibility with an increased focus on programming practices,
transparent reporting and the role of computational environments.
The intended audience is bioinformatics data analysts (not tool developers), however the
principles described here could equally apply to other domains of study.

@@

Containers are an attempt to resolve some of the downsides of VMs.
Container images can be thought of as similar to VMs, but are more lightweight as they
share parts of the host operating system [@Nust2020-cr].
In the example shown in **Figure 4**, the five containerised applications share the same operating
system, while the three VM applications each involve their own operating system which incurs a
significant performance overhead @noauthor_2021-lt.
Therefore, running workflows in containers incurs only a small reduction in performance as compared
to running directly on the host system @Di_Tommaso2015-pw.
Container images are highly portable because containers include everything needed to run the
application, including the system tools and libraries, ensuring that the environment is consistent
across different systems.
For example, the most popular containerisation system, Docker [@Merkel2014-ek], makes it
possible to run Windows and Linux/Unix containers on any computer with Docker installed, with the
promise of reproducibility.
While there are several alternatives such as Podman, Rkt or Apptainer/Singularity, Docker remains the most widely used
containerisation system for web development and research.
Docker now has a large community of users, extensive documentation and
a vast collection of pre-built container images in the DockerHub registry.
Docker can fetch images from DockerHub and run them on the host system with just a couple of commands,
and typically within a few minutes.
This accelerates the installation procedure dramatically which is a known bottleneck for bioinformatic
reproducibility [@Mangul2019-ei;@Mangul2019-xp].

@@

Progress on implementing these best practices into computational research publications has been minimal, as
researchers face a range of obstacles.

These include:

* The "publish-or-perish" incentive structures in academia reward prolific rather than scrupulous
authors.
Part of the problem is the incentive structures in academia, with funding bodies and journals
largely focused on novel research directions rather than confirming previous findings
[@Garijo2013-zl ; @Huang2013-xr; @Kaelin2017-id; @Munafo2017-ty; @Diong2021-is].

* Journal standards are slow to progress.
To illustrate this, many journals still allow publication without deposition of code or data to public
repositories.

* A lack of systematic reproducibility/replication studies in bioinformatics remains small
[@Ioannidis2009-hv; @Zaringhalam2020-ms; @Samuel2022-ha], so the severity of the problem isn't
well understood.

* A lack of documentated protocols and tutorial materials.

* An overall lack of coding/data science expertise among life science researchers and new graduates.

Overcoming these challenges will require a multi-pronged strategy:

>>>>

For example, in 2006 an article entitled "Genomic signatures to guide the use of chemotherapeutics"
generated a great deal of interest as it was an early application of high throughput transcriptomics
in the prediction of individual patient responses to different chemotherapies @Potti2006-yk.
After observing some unusual features of the patient group, Baggerly and Coombes attempted
reproduction of some of the key findings @Baggerly2010-nr.
Without the help of scripted workflows to guide the re-analysis, the team used forensic bioinformatic
techniques to piece together how the dataset was originally analysed.
Their investigations found a litany of issues.
Firstly, the labeling of patients in the test set as "responders" or "non-responders" had been reversed
in the original analysis.
Secondly, some of the patients were included more than once (some up to four times) in the analysis,
likely to cause major distortions in results.
Confusingly, some of the reused data has inconsistent grouping; ie: some of the samples are labeled
both sensitive and resistant.
Additional errors include two cases where results (charts) were ascribed to the wrong drug.
Baggerly and Coombes highlight that such mistakes can inadvertently occur when conducting unscripted
data analysis such as using spreadsheets, and these problems can be obscured by a lack of documentation.
The article underwent two corrigenda, but was untimately retracted in 2011, as the authors were not
able to reproduce the findings themselves due to "corruption of several validation data sets"
@Potti2011-ut.
As the array findings were the basis for clinical trials where patients were allocated to treatments,
the flawed data analysis may have harmed patients given the wrong drug in the
period 2007-2010.
In 2010, Duke University terminated the trials and suspended the lead author, Dr Anil Potti, who
later resigned.
Duke was served eight lawsuits by families of affected patients seeking compensation for exposure to
harmful and unnecessary chemotherapy, which were settled out of court @Dyer2015-my.
This worst-case scenario emphasizes that computational reproducibility is crucial for
translating bioinformatics research into real-world outcomes.

@@

A number of guides recommending enhanced computational reproducibility practices have been developed
[ @Tan2010-is; @Peng2011-ia; @Sandve2013-lk; @Piccolo2016-ji; @Lewis2016-qh;
@Gruning2018-sg; @Akalin2018-eg; @Brito2020-ld; @Peikert2021-eq;
@Rodrigues2023-fw; @DuPre2022-bk; @Krafczyk2021-xv; @Stodden2016-aj].
Inspired by these principles, this article presents the
*five pillars of reproducible computational research* (**Figure 1**).
There are clear parallels with the established three pillars of the open science framework,
(open data, code and papers) [@Gorgolewski2016-hv].
Here, the emphasis is on practical reproducibility with an increased focus on programming practices,
transparent reporting and the role of computational environments.
The intended audience is bioinformatics data analysts (not tool developers), however the
principles described here could equally apply to other domains of study.

@@

Containers are an attempt to resolve some of the downsides of VMs.
Container images can be thought of as similar to VMs, but are more lightweight as they
share parts of the host operating system [@Nust2020-cr].
In the example shown in **Figure 4**, the five containerised applications share the same operating
system, while the three VM applications each involve their own operating system which incurs a
significant performance overhead @noauthor_2021-lt.
Therefore, running workflows in containers incurs only a small reduction in performance as compared
to running directly on the host system @Di_Tommaso2015-pw.
Container images are highly portable because containers include everything needed to run the
application, including the system tools and libraries, ensuring that the environment is consistent
across different systems.
For example, the most popular containerisation system, Docker [@Merkel2014-ek], makes it
possible to run Windows and Linux/Unix containers on any computer with Docker installed, with the
promise of reproducibility.
While there are several alternatives such as Podman, Rkt or Apptainer/Singularity, Docker remains the most widely used
containerisation system for web development and research.
Docker now has a large community of users, extensive documentation and
a vast collection of pre-built container images in the DockerHub registry.
Docker can fetch images from DockerHub and run them on the host system with just a couple of commands,
and typically within a few minutes.
This accelerates the installation procedure dramatically which is a known bottleneck for bioinformatic
reproducibility [@Mangul2019-ei;@Mangul2019-xp;@Krafczyk2021-xv].

@@

Although the technological tools underlying these best practices have existed for a decade,
progress on implementing them into computational research publications has been minimal.

Researchers face several hurdles in adopting these.

* The "publish-or-perish" incentive structures in academia reward prolific rather than scrupulous
authors.
Funding bodies, institutions and journals generally prefer to focus on novel research directions rather
than meticulous confirmation of previous findings [@Garijo2013-zl ; @Huang2013-xr; @Kaelin2017-id;
@Munafo2017-ty; @Diong2021-is].

* A lack of systematic reproducibility/replication studies in bioinformatics remains small
[@Ioannidis2009-hv; @Zaringhalam2020-ms; @Samuel2022-ha], so the severity of the reproducibility
problem isn't well appreciated.

* Journal standards are slow to progress.
Our observations indicate that sharing of data and code remains optional for most medical
journals in PubMed, but in the field of computational biology, providing these resources is
becoming standard practice.

* A lack of documentated protocols and tutorial materials for highly-reproducible analysis makes it
difficult for researchers to confidently update their code development practices.

* An overall lack of coding/data science expertise among life science researchers and new graduates.
Due to strong demand from industry and government, academia has a deficit of workers fluent in data
science [@Committee_on_Envisioning_the_Data_Science_Discipline_The_Undergraduate_Perspective2018-pe].
This makes it challenging for organisations to recruit capable data analysts and cultivate a
community of practice.

Stodden et al have outlined some practical recommendations for tackling these problems [@Stodden2016-aj].

Overcoming these obstacles will require a multi-pronged strategy and effort from different stakeholders.

* What researchers can do.

* What journals can do.

* What insitutions can do.

* What universities can do.

* What funders can do.
